Seagrass mapping and monitoring along the coasts of Crete, Greece

### 2011-02-01 MScDiplom_Lemenkova.tex
changed Document general style (font size and layout), some cosmetic improvements in figures placement. Changed boxed links to colour ones

\documentclass[10pt, a4paper]{article}
\usepackage{color}
\usepackage{setspace}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
\usepackage{makeidx}
\makeindex
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}

@@

\section*{}
	\markright{Table of Content}
	\addcontentsline{toc}{section}{Table of Content}
	\setlength{\textwidth}{6.5in}
	\addtolength{\voffset}{+3pt}
	\textheight = 620pt
	\tableofcontents 
\pagebreak

@@

\section*{Abstract}
	\markright{Summary Abstract of the MSc Thesis}
	\addcontentsline{toc}{section}{Abstract}
The seagrasses, a unique group of aquatic plants, create complex, extremely diversified and productive ecological systems in the littoral coastal zones. The only flowering plant in the world that is able to live completely submerged, seagrasses play vital role in the marine ecosystems of the World Ocean. Seagrasses are the most important component in the environmental food chain of the coastal ecosystems, being a vital food source for various marine species (e.g. fish, dugongs, turtles, swans), and a producer of organic matter, which is the very basis of the food web.\\
The \textit{P.oceanica} seagrass is an endemic for the Mediterranean region, and a main species in the marine coastal environment of Greece. Meadows of \textit{P.oceanica} are subjected to the human activities, because they occur in coastal areas, where they are affected both by anthropogenic and by climatic and environmental factors.  Nowadays \textit{P.oceanica} is in the alarming state of regression, because of the deterioration of the environment in the Mediterranean Sea. Due to these reasons, \textit{P.oceanica} is a protected species since 1988 in some European countries (France). Monitoring \textit{P.oceanica} is therefore an important contribution to the saving and protecting the environment of Mediterranean region.\\
The current MSc thesis focuses on the monitoring of seagrass \textit{P.oceanica} along the northern coasts of Crete Island, Greece, and investigates the application of the remote sensing techniques for the seagrass mapping. 
This research was articulated in two parts, where the first one involves an ecological approach to the seagrass distribution in various regions around the globe and the experience of seagrass monitoring nowadays. The second part of this work has technical character and investigates the application of the remote sensing techniques towards seagrass mapping. It, furthermore, focuses on the optical properties of the \textit{P.oceanica} and other seafloor cover types, and studies distinguishability of various seafloor cover types. Studies of the optical characteristics of separate seafloor cover types were made with purpose to clarify, whether their spectral properties change with varying environmental conditions. \\
Special attention has been drawn on the role of environmental factors on the distribution of \textit{P.oceanica} along the coasts of Crete, and in particular, how the optical properties of the seafloor cover types, i.e. spectral reflectance, are being changed under varying external conditions, e.g. water column, amount of suspended particles and sediments in the seawater, and water temperature. For this purpose we studied differences in the spectral reflectance of \textit{P.oceanica} and other bottom cover types at three distinct depths. The diverse spectral values entail variations in optical properties of the seafloor cover types at changing environmental conditions. We applied WASI simulation techniques for the modelling of the optical parameters of various seafloor cover types by various spaceborne imaging spectrometers (MERIS, SeaWiFS, CZCS and MODIS), in order to understand their suitability and possible limitations for the seagrass mapping.\\
Fieldwork research sites were presented by separate locations on the northern coast of Crete region (Ligaria, Agia Pelagia, Xerocampos). The additional measurements of the reflectance spectra of the seawater with and without sediments have been made in aquarium tank in 2009 by means of Trios-RAMSES spectroradiometer. Parallel to the collection of spectra signatures, we captured the imagery for the seagrass mapping, which consists of the aerial images from the Google Earth website and the satellite Landsat TM and Landsat ETM+ scenes.

@@

\section*{Document Outline}
	\markright{Structure of the MSc Thesis}
	\addcontentsline{toc}{section}{Document Outline}
\textit{Chapter 1}: \nameref{sec:1} is an introductory section. It describes general background of the research problem, outlines the need and actuality of the problem (including the environmental vulnerability of the seagrasses), and highlights limitations and possibilities of the remote sensing application for the current work. In this chapter we also set up research objectives, put research questions for hypothesis testing and sketch research approach for the proposed work.\\
In \textit{Chapter 2}: \nameref{sec:2} we review the existing literature and reported research experience on the similar problem: studies of spectral separability of various seafloor types, limitations and advantages of the remote sensing techniques applied for seagrass mapping seagrass ecology and mapping seagrass environment. We considered not only the Mediterranean environment, but also papers from Australian and Chinese scientists, because seagrass monitoring is most actively developed in the southern regions of our planet. The review of the existing RTM algorithms for the retrieval of the optical parameters, as well as description of various tools for the spectrometric measurements - are given in the same Chapter as well. More close attention has been given to the imaging hyperspectral radiometer Trios RAMSES, used for the data collection.\\
\textit{Chapter 3}: \nameref{sec:3} deals with the methods and materials used in current work. We start our discussion from the fieldwork area location and describe tools and instruments used during the fieldwork, as well as sampling design. Procedures of the data pre-processing and capturing imagery from the Google Earth are also presented in the same chapter. \\
\textit{Chapter 4}: \nameref{sec:4} presents to the reader the main results of the current MSc work, obtained during the processing fieldwork and other collected data. It starts from the review of the collected data from different sources, then describes the modelling component of the research, namely WASI colour simulation and plotted resulting graphs. We also discuss here the particularities of various sensors, suitable enough for deriving radiometric information to discriminate sand from the seagrass. Furthermore, this chapter includes analysis of the spectral signatures of various seafloor cover types in conditions of changing environment, possibility and limitations for \textit{P.oceanica} spectral discrimination, statistical analysis of the data sets and mapping based on images processing.\\
\textit{Chapter 5}: \nameref{sec:5} summarizes and briefly discusses again the main results, reported in detail in a previous chapter. This chapter proposes some final discussions to the reader, and comes to the main research outcomes.\\
Finally, we suggest some recommendations for any further research focusing on the seagrass environment along the coasts of Crete (or other Mediterranean areas) in the final \nameref{sec:6}.\\
A lot of plotted graphs, statistical outcomes, tables, auxiliary (yet relevant to our research) pictures and images, illustrating our work - are collected in the \nameref{app}. Due to the standard editorial limitations of the current MSc work, it was not possible to include all them to the main chapters. However, we tried to make the structure of the Appendices most clear and easy-readable as possible, by dividing the Appendices into 10 various sub-sections and referencing to them from the main text where necessary. \\
The \nameref{bib} section includes literature and internet resources used for the current work. The cross-referencing and web hyper-referencing are used in the whole document to make reading more quick, effective and informative.\\
An Index is placed in the end of the Document to help the reader find what he is looking for.

@@

\section*{List of Acronyms}
	\markright{Acronyms}
	\addcontentsline{toc}{section}{List of Acronyms}
\onehalfspace

\textbf{4AOP}: Automatized Atmospheric Absorption Atlas Operational Release \\
\textbf{6SV1}:  Second Simulation of a Satellite Signal in the Solar Spectrum, Version1\\
\textbf{ANOVA}: ANalysis of VAriance\\
\textbf{ASTER}: Advanced Spaceborne Thermal Emission and Reflection Radiometer\\
\textbf{BRDF}:  Bidirectional Reflectance Distribution Function\\
\textbf{CASI}:		Compact Airborne Spectrographic Imager\\
\textbf{CNES}: 		Centre National d'Etudes Spatiales (of France)\\
\textbf{CRTM}:		Community Radiative Transfer Model\\
\textbf{CZCS}:		 Coastal Zone Color Scanner\\
\textbf{DISORT}:	 Discrete Ordinates Radiative Transfer Program for a Multi-Layered Plane-Parallel Medium\\
\textbf{ENVISAT}:	ENVIronmental SATellite\\
\textbf{ESA}:		European Space Agency\\
\textbf{ERS}: 		European Remote Sensing satellite\\
\textbf{GIS}:		Geographic Information System\\
\textbf{GNU}:		GNU's Not Unix \\
\textbf{GPS}:		 Global Positioning System\\
\textbf{GRASS}:	 Geographic Resources Analysis Support System\\
\textbf{GRETL}:		GNU Regression, Econometrics and Time-series Library\\
\textbf{HyMap}:	Hyperspectral Mapper\\
\textbf{HyperOCR}:	Hyperspectral Ocean Colour Radiometer\\
\textbf{IKONOS}: 	derived its name from the Greek term \textit{eikon} for image\\
\textbf{ILWIS}:		Integrated Land and Water Information System\\
\textbf{iPAQ}:	internet Pocket PC by CompAQ\\
\textbf{KOPRA}:	 Karlsruhe Optimized and Precise Radiative transfer Algorithm\\
\textbf{LaTeX}:		Lamport Tau Epsilon Chi (Greek for \textit{techne}), modified by L.Lamport\\
\textbf{LAI}:	 Leaf Area Index\\
\textbf{Landsat TM}:	 Landsat Thematic Mapper \\
\textbf{Landsat ETM+}:	Landsat Enhanced Thematic Mapper Plus\\
\textbf{LyX}: 		LyriX-files\\
\textbf{MERIS}:		MEdium Resolution Imaging Spectrometer\\
\textbf{MIPAS}:		Michelson Interferometer for Passive Atmospheric Sounding\\
\textbf{MODIS}:	Moderate Resolution Imaging Spectroradiometer\\
\textbf{NASA}:		National Aeronautics and Space Administration\\
\textbf{NIR}:		 Near-Infra-Red\\
\textbf{NOAA}:		 National Oceanic and Atmospheric Administration\\
\textbf{OBC}:		On Board Calibration\\
\textbf{Olympus ST}: 	 Olympus Stylus Tough\\
\textbf{Open EV}: 	Open Earth View\\
\textbf{P.oceanica}:	Posidonia oceanica\\
\textbf{RAMSES}:	RAdiance Measurement SEnSor\\
\textbf{RRTM}:		Rapid Radiative Transfer Model\\
\textbf{RTTOV}: 	Radiative Transfer for TIROS Operational Vertical Sounder\\
\textbf{RTM}:		Radiative Transfer Model\\
\textbf{SAM}:		Scanning Acoustic Microscope\\
\textbf{SCUBA}:	Self Contained Underwater Breathing Apparatus\\
\textbf{SeaWiFS}: 	Sea-viewing Wide Field-of-view Sensor\\
\textbf{SPOT}:		Satellite Pour l'Observation de la Terre\\
\textbf{SPSS	}:	Statistical Package for the Social Sciences\\
\textbf{STDEV}:		Standard Deviation\\
\textbf{SWIR}: 		Short Wave InfraRed\\
\textbf{TIR}: 		Thermal InfraRed\\
\textbf{TIROS}: 	Television and InfraRed Observation Satellite\\
\textbf{UML}:		Unified Modelling Language\\
\textbf{UV}:		UltraViolet (10 nm to 400 nm)\\
\textbf{UVA}:		UltraViolet A, or long waves (315 nm-400 nm)\\
\textbf{UVB}:		 UltraViolet B, or medium waves (280 nm-315 nm)\\
\textbf{VIS}:		VISible\\
\textbf{VNIR}:		Visible and Near-InfraRed\\
\textbf{WASI}:		WAter colour SImulator\\

\pagebreak
\singlespace
\section*{List of Tools}\label{sec:0}
	\markright{List of Tools}
	\addcontentsline{toc}{section}{List of Tools}\index{Software used}

@@

\section*{}
\setcounter{lofdepth}{2}
\markright{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\listoffigures
\pagebreak

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world.\\ Source:\cite{Green03}\label{Green03} 
	\label{fig:2}} 
\end{wrapfigure}

@@

\subsubsection{Ecological significance of the seagrasses}
Seagrass plays vital role in the marine ecosystems of the world ocean. 
\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}
Seagrasses are the only flowering plant in the world that is able to live completely submerged. Seagrass is a habitat for
numerous marine fish species \cite{Nagelkerken00}\label{Nagelkerken00}, source of primary production and food for
fish, turtles and other organisms, which gives them special environmental value \cite{Noralez10}\label{Noralez10}.\\
Seagrass meadows produce enormous quantities of organic matter (leaves, epiphytes), which
constitutes the basis of the food web both within and outside the ecosystem \cite{Gobert06}\label{Gobert06}.
Finally, seagrasses are an important component in the environmental "food chain" of the coastal
ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.\\
Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine “algae”, “seagrasses” are vastly different from them. There are
fundamental differences between both marine organisms, the major of them should be briefly
mentioned: first, seagrasses are true plants with root system and leaves which photosynthesis, have
complex structure and create landscape-similar vast formations on the colonized seafloor areas with
soft sediment, whereas algae are simple organisms that can only holdfast; secondly, seagrasses are
complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while algae
have simple few cell structure with spores and gametes; finally, seagrasses uptake nutrient through
root system, while algae nourish directly from the water column \cite{Dixon05}\label{Dixon05}. There are other
differences between “seagrass” and “algae” but they go beyond the scope of this work.

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}. One
of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.20]{Fig-1-4.jpg}
	\caption{Distribution of seagrass in relation to mean ocean temperature. \\
	Source: \cite{Orth06}\label{Orth06}}
	\label{Fig.4}
\end{wrapfigure}

>>>>

\documentclass[11pt, a4paper]{article}
\usepackage{color}
\usepackage{setspace}
\usepackage[a4paper, left=30mm,right=25mm,top=30mm,bottom=35mm]{geometry}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
\usepackage{makeidx}
\makeindex
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage[a4paper=true, ]{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue}

@@

\section*{}
	\markright{Table of Content}
	\addcontentsline{toc}{section}{Table of Content}
	%\setlength{\textwidth}{6in}
%	\addtolength{\voffset}{+5pt}
%	\marginparwidth = 30pt
%	\textheight = 630pt
	\tableofcontents 
\pagebreak

@@

\section*{Abstract}
	\markright{Summary Abstract of the MSc Thesis}
	\addcontentsline{toc}{section}{Abstract}

The seagrasses, a unique group of aquatic plants, create complex, extremely diversified and productive ecological systems in the littoral coastal zones. The only flowering plant in the world that is able to live completely submerged, seagrasses play vital role in the marine ecosystems of the World Ocean. Seagrasses are the most important component in the environmental food chain of the coastal ecosystems, being a vital food source for various marine species (e.g. fish, dugongs, turtles, swans), and a producer of organic matter, which is the very basis of the food web.\\
The \textit{P.oceanica} seagrass is an endemic for the Mediterranean region, and a main species in the marine coastal environment of Greece. Meadows of \textit{P.oceanica} are subjected to the human activities, because they occur in coastal areas, where they are affected both by anthropogenic and by climatic and environmental factors.  Nowadays \textit{P.oceanica} is in the alarming state of regression, because of the deterioration of the environment in the Mediterranean Sea. Due to these reasons, \textit{P.oceanica} is a protected species since 1988 in some European countries (France). Monitoring \textit{P.oceanica} is therefore an important contribution to the saving and protecting the environment of Mediterranean region.\\
The current MSc thesis focuses on the monitoring of seagrass \textit{P.oceanica} along the northern coasts of Crete Island, Greece, and investigates the application of the remote sensing techniques for the seagrass mapping. 
This research was articulated in two parts, where the first one involves an ecological approach to the seagrass distribution in various regions around the globe and the experience of seagrass monitoring nowadays. The second part of this work has technical character and investigates the application of the remote sensing techniques towards seagrass mapping. It, furthermore, focuses on the optical properties of the \textit{P.oceanica} and other seafloor cover types, and studies distinguishability of various seafloor cover types. Studies of the optical characteristics of separate seafloor cover types were made with purpose to clarify, whether their spectral properties change with varying environmental conditions. \\
Special attention has been drawn on the role of environmental factors on the distribution of \textit{P.oceanica} along the coasts of Crete, and in particular, how the optical properties of the seafloor cover types, i.e. spectral reflectance, are being changed under varying external conditions, e.g. water column, amount of suspended particles and sediments in the seawater, and water temperature. For this purpose we studied differences in the spectral reflectance of \textit{P.oceanica} and other bottom cover types at three distinct depths. The diverse spectral values entail variations in optical properties of the seafloor cover types at changing environmental conditions. We applied WASI simulation techniques for the modelling of the optical parameters of various seafloor cover types by various spaceborne imaging spectrometers (MERIS, SeaWiFS, CZCS and MODIS), in order to understand their suitability and possible limitations for the seagrass mapping.\\
Fieldwork research sites were presented by separate locations on the northern coast of Crete region (Ligaria, Agia Pelagia, Xerocampos). The additional measurements of the reflectance spectra of the seawater with and without sediments have been made in aquarium tank in 2009 by means of Trios-RAMSES spectroradiometer. Parallel to the collection of spectra signatures, we captured the imagery for the seagrass mapping, which consists of the aerial images from the Google Earth website and the satellite Landsat TM and ETM+ scenes.

@@

\section*{Document Outline}
	\markright{Structure of the MSc Thesis}
	\addcontentsline{toc}{section}{Document Outline}
\textit{Chapter 1}: \nameref{sec:1} is an introductory section. It describes general background of the research problem, outlines the need and actuality of the problem (including the environmental vulnerability of the seagrasses), and highlights limitations and possibilities of the remote sensing application for the current work. In this chapter we also set up research objectives, put research questions for hypothesis testing and sketch research approach for the proposed work.
\newline
In \textit{Chapter 2}: \nameref{sec:2} we review the existing literature and reported research experience on the similar problem: studies of spectral separability of various seafloor types, limitations and advantages of the remote sensing techniques applied for seagrass mapping seagrass ecology and mapping seagrass environment. We considered not only the Mediterranean environment, but also papers from Australian and Chinese scientists, because seagrass monitoring is most actively developed in the southern regions of our planet. The review of the existing RTM algorithms for the retrieval of the optical parameters, as well as description of various tools for the spectrometric measurements - are given in the same Chapter as well. More close attention has been given to the imaging hyperspectral radiometer Trios RAMSES, used for the data collection.\\
\textit{Chapter 3}: \nameref{sec:3} deals with the methods and materials used in current work. We start our discussion from the fieldwork area location and describe tools and instruments used during the fieldwork, as well as sampling design. Procedures of the data pre-processing and capturing imagery from the Google Earth are also presented in the same chapter. \\
\textit{Chapter 4}: \nameref{sec:4} presents to the reader the main results of the current MSc work, obtained during the processing fieldwork and other collected data. It starts from the review of the collected data from different sources, then describes the modelling component of the research, namely WASI colour simulation and plotted resulting graphs. We also discuss here the particularities of various sensors, suitable enough for deriving radiometric information to discriminate sand from the seagrass. Furthermore, this chapter includes analysis of the spectral signatures of various seafloor cover types in conditions of changing environment, possibility and limitations for \textit{P.oceanica} spectral discrimination, statistical analysis of the data sets and mapping based on images processing.\\
\textit{Chapter 5}: \nameref{sec:5} summarizes and briefly discusses again the main results, reported in detail in a previous chapter. This chapter proposes some final discussions to the reader, and comes to the main research outcomes.\\
Finally, we suggest some recommendations for any further research focusing on the seagrass environment along the coasts of Crete (or other Mediterranean areas) in the final \nameref{sec:6}.\\
A lot of plotted graphs, statistical outcomes, tables, auxiliary (yet relevant to our research) pictures and images, illustrating our work - are collected in the \nameref{app}. Due to the standard editorial limitations of the current MSc work, it was not possible to include all them to the main chapters. However, we tried to make the structure of the Appendices most clear and easy-readable as possible, by dividing the Appendices into 10 various sub-sections and referencing to them from the main text where necessary. \\
The \nameref{bib} section includes literature and internet resources used for the current work. The cross-referencing and web hyper-referencing are used in the whole document to make reading more quick, effective and informative.\\
An Index is placed in the end of the Document to help the reader find what he is looking for.

@@

\section*{List of Acronyms}
	\markright{Acronyms}
	\addcontentsline{toc}{section}{List of Acronyms}
\doublespacing
\textbf{4AOP}: Automatized Atmospheric Absorption Atlas Operational Release \\
\textbf{6SV1}:  Second Simulation of a Satellite Signal in the Solar Spectrum, Version1\\
\textbf{ANOVA}: ANalysis of VAriance\\
\textbf{ASTER}: Advanced Spaceborne Thermal Emission and Reflection Radiometer\\
\textbf{BRDF}:  Bidirectional Reflectance Distribution Function\\
\textbf{CASI}:		Compact Airborne Spectrographic Imager\\
\textbf{CNES}: 		Centre National d'Etudes Spatiales (of France)\\
\textbf{CRTM}:		Community Radiative Transfer Model\\
\textbf{CZCS}:		 Coastal Zone Color Scanner\\
\textbf{DISORT}:	 Discrete Ordinates Radiative Transfer Program for a Multi-Layered Plane-Parallel Medium\\
\textbf{ENVISAT}:	ENVIronmental SATellite\\
\textbf{ESA}:		European Space Agency\\
\textbf{ERS}: 		European Remote Sensing satellite\\
\textbf{GIS}:		Geographic Information System\\
\textbf{GNU}:		GNU's Not Unix \\
\textbf{GPS}:		 Global Positioning System\\
\textbf{GRASS}:	 Geographic Resources Analysis Support System\\
\textbf{GRETL}:		GNU Regression, Econometrics and Time-series Library\\
\textbf{HyMap}:	Hyperspectral Mapper\\
\textbf{HyperOCR}:	Hyperspectral Ocean Colour Radiometer\\
\textbf{IKONOS}: 	derived its name from the Greek term \textit{eikon} for image\\
\textbf{ILWIS}:		Integrated Land and Water Information System\\
\textbf{iPAQ}:	internet Pocket PC by CompAQ\\
\textbf{KOPRA}:	 Karlsruhe Optimized and Precise Radiative transfer Algorithm\\
\textbf{LaTeX}:		Lamport Tau Epsilon Chi (Greek for \textit{techne}), modified by L.Lamport\\
\textbf{LAI}:	 Leaf Area Index\\
\textbf{Landsat TM}:	 Landsat Thematic Mapper \\
\textbf{Landsat ETM+}:	Landsat Enhanced Thematic Mapper Plus\\
\textbf{LyX}: 		LyriX-files\\
\textbf{MERIS}:		MEdium Resolution Imaging Spectrometer\\
\textbf{MIPAS}:		Michelson Interferometer for Passive Atmospheric Sounding\\
\textbf{MODIS}:	Moderate Resolution Imaging Spectroradiometer\\
\textbf{NASA}:		National Aeronautics and Space Administration\\
\textbf{NIR}:		 Near-Infra-Red\\
\textbf{NOAA}:		 National Oceanic and Atmospheric Administration\\
\textbf{OBC}:		On Board Calibration\\
\textbf{Olympus ST}: 	 Olympus Stylus Tough\\
\textbf{Open EV}: 	Open Earth View\\
\textbf{RAMSES}:	RAdiance Measurement SEnSor\\
\textbf{RRTM}:		Rapid Radiative Transfer Model\\
\textbf{RTTOV}: 	Radiative Transfer for TIROS Operational Vertical Sounder\\
\textbf{RTM}:		Radiative Transfer Model\\
\textbf{SAM}:		Scanning Acoustic Microscope\\
\textbf{SCUBA}:	Self Contained Underwater Breathing Apparatus\\
\textbf{SeaWiFS}: 	Sea-viewing Wide Field-of-view Sensor\\
\textbf{SPOT}:		Satellite Pour l'Observation de la Terre\\
\textbf{SPSS	}:	Statistical Package for the Social Sciences\\
\textbf{STDEV}:		Standard Deviation\\
\textbf{SWIR}: 		Short Wave InfraRed\\
\textbf{TIR}: 		Thermal InfraRed\\
\textbf{TIROS}: 	Television and InfraRed Observation Satellite\\
\textbf{UML}:		Unified Modelling Language\\
\textbf{UV}:		UltraViolet (10 nm to 400 nm)\\
\textbf{UVA}:		UltraViolet A, or long waves (315 nm-400 nm)\\
\textbf{UVB}:		 UltraViolet B, or medium waves (280 nm-315 nm)\\
\textbf{VIS}:		VISible\\
\textbf{VNIR}:		Visible and Near-InfraRed\\
\textbf{WASI}:		WAter colour SImulator
\pagebreak
\singlespace
\section*{List of Tools}\label{sec:0}
	\markright{List of Tools}
	\addcontentsline{toc}{section}{List of Tools}\index{Software used}

@@

\section*{}
\setcounter{lofdepth}{2}
\markright{List of Figures}
\addcontentsline{toc}{section}{List of Figures}
\onehalfspacing
\listoffigures
\pagebreak

@@

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.30]{Fig-1-2.jpg}
	\caption{Distribution of seagrasses in the world.\\ Source:\cite{Green03}\label{Green03} 
	\label{fig:2}} 
\end{wrapfigure}

@@

\subsubsection{Ecological significance of the seagrasses}
Seagrass plays vital role in the marine ecosystems of the world ocean. 
Seagrasses are the only flowering plant in the world that is able to live completely submerged. Seagrass is a habitat for
numerous marine fish species \cite{Nagelkerken00}\label{Nagelkerken00}, source of primary production and food for
fish, turtles and other organisms, which gives them special environmental value \cite{Noralez10}\label{Noralez10}.\\
Seagrass meadows produce enormous quantities of organic matter (leaves, epiphytes), which
constitutes the basis of the food web both within and outside the ecosystem \cite{Gobert06}\label{Gobert06}.
Finally, seagrasses are an important component in the environmental "food chain" of the coastal
ecosystems, being the food source for dugongs, turtles, swans and various fish \cite{Cappo95}\label{Cappo95}.

\begin{wrapfigure}{r}{0.4\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-3.jpg}
	\caption{Morphology of different types of seagrasses. Source: \url{http://www.ozcoasts.org.au}}
	\label{fig:3}
\end{wrapfigure}

Due to their wide distribution, meadows size, easy collection and abundance, sensitivity to the
modifications of the coastal zone and their important role in maintaining coastal water quality
and clarity, seagrass is perfect indicator and descriptor of the environmental health of marine
ecosystems, and is highly suitable for the environmental monitoring \cite{Pergent-Martini05}\label{Pergent-Martini05}.\\
Being often confused with marine “algae”, “seagrasses” are vastly different from them. There are
fundamental differences between both marine organisms, the major of them should be briefly
mentioned: first, seagrasses are true plants with root system and leaves which photosynthesis, have
complex structure and create landscape-similar vast formations on the colonized seafloor areas with
soft sediment, whereas algae are simple organisms that can only holdfast; secondly, seagrasses are
complex vascular plants with reproductive mechanism such as fruits, seeds and spores, while algae
have simple few cell structure with spores and gametes; finally, seagrasses uptake nutrient through
root system, while algae nourish directly from the water column \cite{Dixon05}\label{Dixon05}. There are other
differences between “seagrass” and “algae” but they go beyond the scope of this work.

\subsubsection{Environmental vulnerability of the seagrasses}
Meadows of \textit{P. oceanica} are subjected to the human activities\index{Seagrasses!ecological vulnerability}, as they occur in coastal areas, where
they can be affected both directly \cite{Meinesz91}\label{Meinesz91} or indirectly, through the impact on the
quality of waters and sediments \cite{Duarte02}\label{Duarte02}. As \textit{P.oceanica} is a long-living plant with a slow
growth rate, the anthropogenic modifications of the coastal zone, happening more rapidly than the
capacity of the plant to adapt to these changes, reduce its distribution area \cite{Micheli05}\label{Micheli05}.

\begin{wrapfigure}{r}{0.5\textwidth}
	\centering
	\includegraphics[scale=0.25]{Fig-1-4.jpg}
	\caption{Distribution of seagrass in relation to mean ocean temperature. \\
	Source: \cite{Orth06}\label{Orth06}}
	\label{Fig.4}
\end{wrapfigure}

One of the main drivers of seagrass decline is, for instance, the location of the fish farming near the
seagrass meadows. The negative effects of the sedimentation of waste particles in the farm vicinities
on \textit{P.oceanica} meadows are diverse and complex, and may cause benthic deterioration, accumulation
of organic matter and seagrass decline \cite{Holmer08}\label{Holmer08}. Seagrasses are subject to anthropogenic
nutrient (N and P) loading, which may occasionally casue morphological (e.g. leaf length) and
physiological (e.g. chlorophyll and nitrogen content of the leaves) responses towards changed
environmental conditions \cite{Leoni06,Leoni07}\label{Leoni06} \label{Leoni07}. 
The detailed research of the fish
farm-induced decline of the seagrass meadows \cite{Diaz-Almela06}\label{Diaz-Almela06}reports the relationships of
fish farm organic and nutrient content in the sediments with dynamics of the key seagrass species (\textit{P.
oceanica}) in the Mediterranean Sea. Nowadays \textit{P.oceanica} is in the alarming state of regression due
to the deterioration of the environment in the Mediterranean Sea \cite{Ribed02}\label{Ribed02}. Due to these reasons,
\textit{P.oceanica} is a protected species since 1988 in some European countries \cite{Francour99}\label{Francour99}, and
its presence serves as an indicator of a stable healthy environment. Among other negative factors,
affecting both growth and status of the seagrasses the environmental contaminants can be mentioned,
e.g. thermal, sewage, dredging and chemical pollution\index{Seagrasses!anthropogenic impacts on}  as well as any other kind of maritime works,
e.g. trawling and anchoring of boats \cite{Ribed02}\label{Ribed02}. Other human activities that cause degrading of the
seagrass are recreational boating, commercial overexploitation of coastal resources, eutrophication
\cite{McKenzie06}\label{McKenzie06}.\\
Besides anthropogenic factors, various biochemical, climatic and environmental processes\index{Seagrasses!natural impacts on} can cause
negative influence on seagrass distribution. Seagrass is exposed to threats from the global climate and
environmental change, i.e. increases in sea surface temperature; sea level rise; increased frequency
and intensity of storms and waves; local decrease of water quality, increased sedimentation,
contamination and nitrification; desiccation; salinity fluctuations; nutrient changes; suspended
sediments \cite{Blake00}\label{Blake00}. These stress-drivers can alone result in large-scale seagrass
degradation, but often seagrass undergo simultaneous affects from several of these factors together.
It naturally increases the environmental pressing  to drastic loss of very large areas of
seagrass globally \cite{Orth06}\label{Orth06}.
In tropical areas, where most of seagrasses are located (Fig.\ref{Fig.4}), seagrasses are subjectto catastrophic
extinction and loss, due to the cyclones, typhoons, storms, regular floods and increased rainfalls. Recovery from such
events can take up to several years and often it is only possible by means of the seed reserves from the local
environmental surveys \cite{McKenzie07}\label{McKenzie07}. Other threats for seagrasses in tropical areas are 
increased nutrient availability in the coastal zones, increased
eutrophication and invasive macroalgae. These processes have strong affect on the status of the
seagrass meadows, and often lead to their complete disappearance \cite{Holmer09}\label{Holmer09}.\\
Other environmental threat for the seagrasses arises as a result of the environmental struggle and
competition for existence among species. Thus, meadows of P.oceanica in the Mediterranean Sea are
presently facing invasion by alien algal species, particularly in areas where \textit{P.oceanica} is already
degrading, stressed, have gaps and patchy structure in meadows and show other signs of regression
\cite{Montefalcone10}\label{Montefalcone10}.
Seagrasses are vulnerable fragile species, important for the marine coastal ecosystems, especially for
the protection of the beach structure. However, the facts about seagrass global degrading sound
worrying: about 54 \% of the total seagrass meadows have lost any part of their area; the areas, where
